\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity}

### 2013-12-09 docs/paper/paper.tex
Hier en daar wat geschaafd aan tekst.

We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or a standard approach such as the PC-algorithm \cite{spirtes2000}, named after \textbf{P}eter Spirtes and \textbf{C}lark Glymour, can solve some of the difficulties effective connectivity suffers from.

@@

\subsection{Causal discovery}
An approach for finding effective connectivity may be found in the domain of causal discovery.
A causal discovery method uses data about events to determine which of these events are correlated, and if possible, it determines which event causes which.
The definition of a causal relation is not trivial \cite{} and the relevant principles will be covered in this section.
In our context, the causal discovery methods which have been used aim at inferring a Bayesian network.
This is done by calculating statistical (in)dependencies between stochastic variables.

Such a Bayesian network can be modelled as a graph with edges which may be directed or undirected.
Each variable is represented by a node, and correlations between variables are represented by edges between those nodes.
If there is an edge between two variables $X$ and $Y$, this is denoted as $X - Y$.
A causal relation is denoted with directed edges; $X \rightarrow Y$ implies that $X$ is a direct cause of $Y$.
It is also said that $X$ is the parent of $Y$, and $Y$ is the child of $X$. 
When there is a directed path $X \rightarrow Z_1 \rightarrow Z_2 \dots \rightarrow Y$, $X$ is said to be an ancestor of $Y$ and $Y$ is a descendant of $X$.
An ancestor is an (possibly indirect) cause of its descendant.
A Bayesian network cannot contain cycles, as that would imply that a variable is it's own ancestor and thus it's own cause.
Ideally, a causal discovery methods finds only directed edges, resulting in a directed acyclic graph (DAG).
In practice, however, often not all causal relations can be inferred and undirected edges are also used in a graph.
As such, the result of a causal discovery method is a partially directed acyclic graph (PDAG).
Image {plaatje!} shows an example of a Bayesian network.
A causal discovery method attempts at finding a real-world Bayesian network.
But real-world situations like a brain are immensely complex and contain an unimaginable large number of atoms, which could all function as seperate variables.
As such, we represent our real-world, underlying Bayesian network with a much simpler model.
Such a model contains a reduced number of variables to be able to analyse and visualize the data.

@@

We did not perform explicit (pre-)processing on the functional and diffusion imaging data.
Instead, we used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as (pre-)processed by Hinne et al. \cite{hinne2013}.

@@

One notable difference between the results found in Hinne et al. are that connections between the the two hemispheres are clearly visible as two diagonal lines. Connections within the hemispheres are less clearly visible, as well as connections within the cerebellum.

@@

\section{Discussion}

@@

>>>>

We would like to know whether applying such methods, such as the approach introduced in \cite{claassen2012}, or a standard approach such as the PC-algorithm \cite{spirtes2000}, named after its authors \textbf{P}eter Spirtes and \textbf{C}lark Glymour, can solve some of the difficulties effective connectivity suffers from.

@@

\subsection{Causal discovery}
An approach for finding effective connectivity may be found in the domain of causal discovery.
A causal discovery method uses data about events to determine which of these events are correlated, and additionally, it attempts to determine which events cause which.
The definition of a causal relation is not trivial \cite[p.20]{spirtes2000} and the relevant principles will be covered in this section.
In our context, the causal discovery methods which have been used aim at inferring a Bayesian network or Directed Acyclic Graph (DAG).
These can be interpreted from a probabilistic and a causal perspective.
In our case, we consider graphs that also include non-directed edges, a so called Partial DAG (PDAG).

Under the causal interpretation, a PDAG $G$ each variable is represented by a node and correlations between variables are represented by edges between those nodes.
If such an edge exists between two variables $X$ and $Y$, this is denoted as $X - Y$.
A causal relation is denoted with directed edges; $X \rightarrow Y$ implies that $X$ is a direct cause of $Y$.
It is also said that $X$ is the parent of $Y$, and $Y$ is the child of $X$. 
When there is a directed path $X \rightarrow Z_1 \rightarrow Z_2 \dots \rightarrow Y$, $X$ is said to be an ancestor of $Y$ and $Y$ is a descendant of $X$.
An ancestor is an (possibly indirect) cause of its descendant.

% TODO: Add example of a Bayesian network.
%Image {plaatje!} shows an example of a Bayesian network.

A causal discovery method attempts at finding a real-world Bayesian network.
The human brain is immensely complex and contains a computationally intractable amount of neurons to model.
As such, we represent our real-world, underlying Bayesian network with a much simpler model.
Such a model contains a reduced number of variables in order to remain tractable and to visually interpret model results.

@@

We did not perform explicit (pre-)processing on the functional and diffusion imaging data.
Instead, we used the six resulting resting-state functional time series data and connectivity matrices from the Bayesian analysis as (pre-)processed by Hinne et al. \cite{hinne2013}.
% K is de MAP estimate uit een G-Wishart distributie. Dat wil zoveel zeggen als: een distributie over alle precisiematrices die nullen hebben op de plekken waar G een 0 heeft (Hastie, Trevor; Tibshirani, Robert; Friedman, J. (2009). The Elements of Statistical Learning, voor de ge•nteresseerden).

@@

One notable difference between the results found in Hinne et al. are the strong connections between homological areas, visible as two diagonal lines.
This is a known shortcoming of diffusion weighted imaging, as these connections are typically too long to measure correctly.
These connections most likely correspond to the corpus callosum.
Connections within the hemispheres are less clearly visible, as well as connections within the cerebellum.

@@

\section{Discussion}
% Further research
A Bayesian network cannot contain cycles, as that would imply that a variable is it's own ancestor and thus it's own cause.
The theory used here does not incorporate a temporal component and 
Ideally, a causal discovery methods finds only directed edges, resulting in a directed acyclic graph (DAG).
%In practice, however, often not all causal relations can be inferred and undirected edges are also used in a graph.
%As such, the result of a causal discovery method is a partially directed acyclic graph (PDAG).
As contemporary measuring techniques are too coarse and slow to model a brain at one point in time, correctly modelling variables over multiple points in time - opposing the acyclic assumption - would provide an interesting direction of further research.

The

@@

\section{Acknowledgements}
We would like to thank Max Hinne and Tom Claassen, for without them this research would not have been viable.
They have provided us with ideas, valuable discussion and the data to perform actual research.
